\section{Objectives \& Contributions}
\label{sec:objectives}

This thesis aims to extend existing iterative refinement approaches to hierarchical low-rank matrices. Since iterative refinement in three precisions does not require accurate LU factors, the key idea is to replace the low precision calculations with hierarchical low-rank approximations. These formats have the advantage of being able to calculate a LU factorization at a reduced complexity of $\mathcal{O}(n\;log(n))$ while maintaining a controllable degree of accuracy. By adjusting the error threshold of the low-rank factorization, the same error bounds as for the low precision case should be applicable in theory. Verifying this hypothesis while also providing a first performance benchmark on the combination of those two techniques is the main goal of this research.

Additionally,  we hope to provide some insights into the problem itself via a detailed analysis of the  proposed framework. This includes general convergence of the algorithm as well as the performance trade-off between accuracy of the approximation and number of iterative refinement steps necessary. Since the efficiency of hierarchical representations is tied to is tied to the maintained accuracy, it should be as low as possible while still being able to deliver accurate results. Hence, investigating this relationship is an integral part of this research. 

Finally, experimental results should be obtained to determine future areas of research. This includes the combination of low precision and low-rank approximations (i.e. an approximate factorization in lower precision) as well as efforts to explore parallelism for the algorithm. Even though this examination is limited to single, double and quadruple precisions \cite{institute_of_electrical_and_electronics_engineers_ieee_2008}, due to restricted support from some of the libraries used, an extension to half precision is planned as part of the future work.

By keeping the above-mentioned objectives in mind, this thesis was able to achieve to following contributions to the field of iterative refinement:
\begin{itemize}
    \item Establishing a first performance benchmark for iterative refinement in three precisions based on a comparison to LAPACK routines.
    \item A detailed error analysis of hierarchical low-rank iterative refinement, demonstrating its abilities to obtain accurate results even if the approximation accuracy is less than half the working precision.
    \item Reducing the runtime complexity of iterative refinement to $\mathcal{O}(n^2)$ due to the hierarchical factorization.
    \item A first analysis of low-precision low-rank iterative refinement, showing that the error bound are still applicable while the iteration count increases.
    \item Demonstrating that cost of quadruple precision calculations is generally too large to be considered useful, as it offsets the potential gains from the approximate factorization.
    \item Providing some first insights into parallelization techniques for iterative refinement.
\end{itemize}

